Title Text

### 2011-06-09 talk/iwtc11/paper.tex
stolen example from previous paper and adapted it for our needs. the description of the example probably also needs to be adapted...

\documentclass{sigplanconf}

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

\usepackage{listings}

\usepackage[T1]{fontenc}
\usepackage{beramono}

\definecolor{gray}{rgb}{0.3,0.3,0.3}

\lstset{
  basicstyle=\setstretch{1.05}\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{gray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
  numberstyle = \tiny,
  numbersep = -20pt,
}

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}
\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}

\newcommand\ie{i.e.,\xspace}
\newcommand\eg{e.g.,\xspace}
\newcommand{\etal}{\emph{et al.}\xspace}

\normalem

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

%
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%
\begin{document}
\conferenceinfo{PEPM'11,} {January 24--25, 2011, Austin, Texas, USA.}
\CopyrightYear{2011}
\copyrightdata{978-1-4503-0485-6/11/01}

\title{Allocation Removal by Partial Evaluation in a Tracing JIT}

\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany

            $^b$merlinux GmbH, Hildesheim, Germany

            $^c$Open End, Göteborg, Sweden
           }
           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}

%\numberofauthors{3}
%\author{
%\alignauthor Carl Friedrich Bolz\\
%       \email{cfbolz@gmx.de}
%\alignauthor Michael Leuschel\\
%       \email{leuschel@cs.uni-duesseldorf.de}
%\alignauthor David Schneider\\
%      \email{david.schneider@uni-duesseldorf.de}
%      \sharedaffiliation
%      \affaddr{Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany}\\
%}

\maketitle
\begin{abstract}
The performance of many dynamic language implementations suffers from
high allocation rates and runtime type checks.  This makes dynamic
languages less applicable to purely algorithmic problems, despite their
growing popularity.  In this paper we present a simple compiler optimization
based on online partial evaluation to remove object allocations and
runtime type checks in the context of a tracing JIT.  We evaluate the
optimization using a Python VM and find that it gives good results for
all our (real-life) benchmarks.
\footnote{This research is partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
Eureka Eurostars).}
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\category{D.3.4}{Programming Languages}{Processors}[code generation,
interpreters, run-time environments]

\terms
Languages, Performance, Experimentation

\keywords{Tracing JIT, Partial Evaluation, Optimization}

@@

The objective of a just-in-time (JIT) compiler for a dynamic language is to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is therefore to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

Boxing of primitive types is necessary because dynamic languages need to be able to handle
all objects, even integers, floats, booleans etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, \ie a small
heap-structure is allocated for them that contains the actual value. Boxing
primitive types can be very costly, because a lot of common operations,
particularly all arithmetic operations, have to produce new boxes, in addition
to the actual computation they do. Because the boxes are allocated on the heap,
producing many of them puts pressure on the garbage collector.

Type dispatching is the process of finding the concrete implementation that is
applicable to the objects at hand when performing a generic operation on them. An
example would be the addition of two objects: For addition the types of the
concrete objects need to be checked and the suiting implementation chosen.
Type dispatching is a very common operation in
modern\footnote{For languages in the LISP family, basic arithmetic operations
are typically not overloaded; even in Smalltalk, type dispatching is much
simpler than in Python or JavaScript.}
dynamic languages because no types are known at compile time. Therefore all
operations need it.

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its commonly executed parts into \emph{linear execution traces}. Those
traces are optimized and turned into machine code.

One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an existing interpreter, reusing a lot of
the interpreter's infrastructure. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many of the hard algorithms in
a compiler, such as register allocation.

The use of a tracing JIT can remove the overhead of bytecode dispatch and that
of the interpreter data structures. In this paper we want to present a new
optimization that can be added to a tracing JIT that further removes some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. Our experimental platform is the PyPy project, which is an
environment for implementing dynamic programming languages. PyPy and tracing
JITs are described in more detail in Section~\ref{sec:Background}.
Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
objects that are allocated in the trace \emph{static,} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}; a more formal description is
given in Section~\ref{sec:formal}. The introduced
techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
interpreter.

The contributions made by this paper are:

\begin{enumerate}
    \item A description of a practical, efficient and effective algorithm for
    removing object allocations in a tracing JIT.
    \item A characterization of this algorithm as partial evaluation.
    \item Performance benchmarks for this algorithm.
\end{enumerate}

\section{Background}
\label{sec:Background}

\subsection{PyPy}
\label{sub:PyPy}

The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on C level. During translation to C, many low-level
aspects of the final VM, such as object layout, garbage collection and memory
model, are woven into the generated code. Therefore the interpreter itself can
remain at a relatively high level of abstraction.

A number of languages have been implemented with PyPy. The project was initiated
to get a better Python implementation, which inspired the name of the project
and is still the main focus of development. In addition a number of other
languages were implemented, among them a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010}, a Smalltalk VM
\cite{carl_friedrich_bolz_back_2008} and a GameBoy emulator
\cite{bruni_pygirl:_2009}.

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a tracing just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer using a small number of
source-code hints in the interpreter. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Optimizations can be shared between different languages implemented with PyPy.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code.  Examples of
JIT backends that are implemented are those for Intel x86 and x86-64 and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.

\subsection{Tracing JIT Compilers}
\label{sub:JIT_background}

Tracing JITs are a recently popular approach to write just-in-time compilers for
dynamic languages. Their origins lie in the Dynamo project, which used a tracing
approach to optimize machine code using execution traces
\cite{bala_dynamo:_2000}. Tracing JITs have then be adapted to be used for a
very light-weight Java VM \cite{gal_hotpathvm:_2006} and afterwards used in
several implementations of dynamic languages, such as JavaScript
\cite{andreas_gal_trace-based_2009}, Lua\footnote{\texttt{http://luajit.org/}}
and now Python (and other languages) via PyPy.

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the commonly executed, \ie \emph{hot} paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
mostly mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.
This trace of operations subsequently forms the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation are simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through the loop.
To make sure that the trace maintains the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an if-statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the if-statement is the same as
during tracing, because if
it isn't, the rest of the trace would not be valid.

When generating machine code, every guard is turned into a quick check to
see whether the assumption still holds. When such a guard is hit during the
execution of the machine code and the assumption does not hold, the execution of
the machine code is stopped, and interpreter continues to run from that point
on. These guards are the only mechanism to stop the execution of a trace, the
loop end condition also takes the form of a guard.

If one specific guard fails a lot (\ie more than some threshold), the tracing JIT will generate a new
trace that starts exactly at the position of the failing guard
\cite{andreas_gal_incremental_2006}. The existing assembler is patched to jump
to the new trace when the guard fails.  This approach guarantees that all the
hot paths in the program will eventually be traced and compiled into efficient
code.

@@

def f(y):
   res = BoxedInteger(0)
   while y.is_positive():
      res = res.add(y).add(BoxedInteger(-100))
      y = y.add(BoxedInteger(-1))
   return res
\end{lstlisting}
\caption{An ``Interpreter'' for a Tiny Dynamic Language Written in RPython}
\label{fig:objmodel}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
# arguments to the trace: $p_{0}$, $p_{1}$
# inside f: res.add(y)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)

# inside f: BoxedInteger(-100) 
$p_{6}$ = new(BoxedInteger)
    # inside BoxedInteger.__init__
    set($p_{6}$, intval, -100)

# inside f: .add(BoxedInteger(-100))
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{7}$ = get($p_{5}$, intval)
    guard_class($p_{6}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{8}$ = get($p_{6}$, intval)
        $i_{9}$ = int_add($i_{7}$, $i_{8}$)
        $p_{10}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{10}$, intval, $i_{9}$)

# inside f: BoxedInteger(-1)
$p_{11}$ = new(BoxedInteger)
    # inside BoxedInteger.__init__
    set($p_{11}$, intval, -1)

# inside f: y.add(BoxedInteger(-1))
guard_class($p_{0}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{12}$ = get($p_{0}$, intval)
    guard_class($p_{11}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{13}$ = get($p_{11}$, intval)
        $i_{14}$ = int_add($i_{12}$, $i_{13}$)
        $p_{15}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{15}$, intval, $i_{14}$)

# inside f: y.is_positive()
guard_class($p_{15}$, BoxedInteger)
    # inside BoxedInteger.is_positive
    $i_{16}$ = get($p_{15}$, intval)
    $i_{17}$ = int_gt($i_{16}$, 0)
# inside f
guard_true($i_{17}$)
jump($p_{15}$, $p_{10}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

\section{Object Lifetimes in a Tracing JIT}
\label{sec:lifetimes}

% section Object Lifetimes in a Tracing JIT (end)

To understand the problems that this paper is trying to solve in more detail, we
first need to understand various cases of object lifetimes that can occur in a
tracing JIT compiler.

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{figures/obj-lifetime}
\end{center}

\caption{Object Lifetimes in a Trace}
\label{fig:lifetimes}
\end{figure}

Figure~\ref{fig:lifetimes} shows a trace before optimization, together with the
lifetime of various kinds of objects created in the trace. It is executed from
top to bottom. At the bottom, a jump is used to execute the same loop another
time (for clarity, the figure shows two iterations of the loop). The loop is
executed until one of the guards in the trace fails, and the execution is
aborted and interpretation resumes.

Some of the operations within this trace are \lstinline{new} operations, which each
create a new instance of some class. These instances are used for some time, \eg
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a
non-inlined function via a residual call.

Together with the \lstinline{new} operations, the figure shows the lifetimes of the
created objects. The objects that are created within a trace using \lstinline{new}
fall into one of several categories:

\begin{enumerate}
    \item Objects that live for some time, and are then just not
    used any more afterwards.

    \item Objects that live for some time and then escape.

    \item Objects that live for some time, survive across the jump to
    the beginning of the loop, and are then not used any more.

    \item Objects that live for some time, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.
\end{enumerate}

The objects that are allocated in the example trace in
Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
$p_{5}$, $p_{6}$, $p_{11}$ are in category 1, objects in $p_{10}$, $p_{15}$ are in
category 3.

The creation of objects in category 1 is removed by the optimization described
in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
categories are partially optimized by this approach as well.\footnote{We also started to
work on optimizing objects in category 3, which will be the subject of a later
paper.}

\section{Allocation Removal in Traces}
\label{sec:statics}

The main insight to improve the code shown in Section~\ref{sub:example} is that objects
in category 1 do not survive very long -- they are used only inside the loop and
there is no other outside reference to them. The optimizer identifies objects in
category 1 and removes the allocation of these objects, and all operations
manipulating them.

This is a process that is usually called \emph{escape analysis}
\cite{goldberg_higher_1990}. In this paper we will perform escape analysis by
using partial evaluation. The use of partial evaluation is a bit peculiar in
that it receives no static input arguments for the trace, but it is only used to
optimize operations within the trace. This section will give an informal account
of this process by examining the example trace in Figure~\ref{fig:unopt-trace}.
The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
line numbers are the lines of the unoptimized trace where the operation originates).

To optimize the trace, it is traversed from beginning to end and an output
trace is produced. Every operation in the input trace is either
removed or copied into the output trace. Sometimes new operations need to be
produced as well. The optimizer can only remove operations that manipulate
objects that have been allocated within the trace, while all other operations are copied to the
output trace unchanged.

Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
in lines 1--9 are manipulating objects which existed before the trace and that
are passed in as arguments: therefore the optimizer just copies them into the
output trace.

The following operations (lines 10--17) are more interesting:

\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$p_{5}$ = new(BoxedInteger)   |\setcounter{lstnumber}{11}|
set($p_{5}$, intval, $i_{4}$) |\setcounter{lstnumber}{14}|
$p_{6}$ = new(BoxedInteger)   |\setcounter{lstnumber}{16}|
set($p_{6}$, intval, -100)
\end{lstlisting}

When the optimizer encounters a \lstinline{new}, it removes it optimistically,
and assumes that the object is in category 1. If later the optimizer finds that the object escapes, it
will be allocated at that point. The optimizer needs to keep track
of the state of the object that the operation would have created.
This is done with the help of a \emph{static object}\footnote{Here
``static'' is meant in the sense of partial evaluation, \ie known at partial
evaluation time, not in the sense of ``static allocation'' or ``static
method''.}. The static object describes the shape of the object that would have
been allocated, \ie the type of the object and where the values that would be
stored in the fields of the allocated object come from.

In the snippet above, the two \lstinline{new} operations are removed and two
static objects are constructed. The \lstinline{set} operations manipulate
static objects, therefore they can be removed as well; their effect is
remembered in the static objects.

After the operations the static object associated with $p_{5}$ would store the knowledge that it is a
\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would store that it is a \lstinline{BoxedInteger}
whose \lstinline{intval} field contains the constant -100.

The subsequent operations (line 20--26) in Figure~\ref{fig:unopt-trace}, which
use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:

\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
guard_class($p_{5}$, BoxedInteger)  |\setcounter{lstnumber}{21}|
$i_{7}$ = get($p_{5}$, intval)
guard_class($p_{6}$, BoxedInteger)  |\setcounter{lstnumber}{24}|
$i_{8}$ = get($p_{6}$, intval)
$i_{9}$ = int_add($i_{7}$, $i_{8}$)
\end{lstlisting}

The \lstinline{guard_class} operations can be removed, since their arguments are
static objects with the matching type \lstinline{BoxedInteger}. The
\lstinline{get} operations can be removed as well, because each of them reads a
field out of a static object. The results of the get operation are replaced with
what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
replaced by $i_{4}$ and -100. The only operation copied into the optimized trace
is the addition:

\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$i_{9}$ = int_add($i_{4}$, -100)
\end{lstlisting}

The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
similar vein. The operations in lines 27--35 produce two more static objects and
are removed. Those in line 36--39 are just copied into the output trace because they
manipulate objects that are allocated before the trace. Lines 40--42 are removed
because they operate on a static object. Line 43 is copied into the output trace.
Lines 44--46 produce a new static object and are removed, lines 48--51 manipulate
that static object and are removed as well. Lines 52--54 are copied into the output
trace.

The last operation (line 55) is an interesting case. It is the \lstinline{jump}
operation that passes control back to the beginning of the trace. The two
arguments to this operation at this point are static objects. However, because
they are passed into the next iteration of the loop they live longer than the trace
and therefore cannot remain static. They need to be turned into dynamic
(runtime) objects before the actual \lstinline{jump} operation. This process of
turning a static object into a dynamic one is called \emph{lifting}.

Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
the output trace. Those operations produce an object at runtime that has the
shape described by the static object. This process is a bit delicate,
because the static objects could form an arbitrary graph structure. In our
example it is simple, though:

\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{54}|
jump($p_{15}$, $p_{10}$)
\end{lstlisting}

Observe how the operations for creating these two instances have been moved to a
later point in the trace. This is worthwhile even though the objects have to be
allocated in the end because some \lstinline{get} operations and
\lstinline{guard_class} operations on the lifted static objects could be
removed.

More generally, lifting needs to occur if a static object is used in any
operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
It also needs to occur if \lstinline{set} is used to store a static object into
a non-static one.

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original
five, and only three \lstinline{guard_class} operations, compared to the
original seven.

\begin{figure}
\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
guard_class($p_1$, BoxedInteger)           |\setcounter{lstnumber}{4}|
$i_2$ = get($p_1$, intval)
guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{7}|
$i_3$ = get($p_0$, intval)
$i_4$ = int_add($i_2$, $i_3$)              |\setcounter{lstnumber}{25}|
$i_9$ = int_add($i_4$, -100)               |\setcounter{lstnumber}{35}|

guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{38}|
$i_{12}$ = get($p_0$, intval)              |\setcounter{lstnumber}{42}|
$i_{14}$ = int_add($i_{12}$, -1)           |\setcounter{lstnumber}{50}|

$i_{17}$ = int_gt($i_{14}$, 0)             |\setcounter{lstnumber}{53}|
guard_true($i_{17}$)                       |\setcounter{lstnumber}{42}|

$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{53}|

jump($p_{15}$, $p_{10}$)
\end{lstlisting}

\caption{Resulting Trace After Allocation Removal}
\label{fig:step1}
\end{figure}

\section{Formal Description of the Algorithm}
\label{sec:formal}

\begin{figure*}
\begin{center}
\begin{tabular}{lcccc}
\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathtt{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ & \emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{\,}{u=\mathtt{get}(v,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[u\mapsto H\left(E\left(v\right)\right)_{F}\right],H}}$ & ~~~ &  & ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{\,}{\mathtt{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &  & \tabularnewline[4em]
\end{tabular}
\end{center}

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\caption{The Operational Semantics of Simplified Traces}
\label{fig:semantics}
\end{figure*}

In this section we want to give a formal description of the semantics of the
traces and of the optimizer and liken the optimization to partial evaluation.
We focus on the operations for manipulating heap allocated objects,
as those are the only ones that are actually optimized. We also consider only
objects with two fields $L$ and $R$ in this section, generalizing to arbitrary many fields
is straightforward. Traces are lists of operations. The operations considered here are
\lstinline{new}, \lstinline{get}, \lstinline{set} and \lstinline{guard_class}.

The values of all
variables are locations (\ie pointers). Locations are mapped to objects, which
are represented by triples $(T,l_1,l_2)$ of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are initialized to a real
location before being read, otherwise the trace is malformed (this condition is
guaranteed by how the traces are generated in PyPy).

We use some abbreviations when dealing with object triples. To read the type of
an object, $\mathrm{type}((T,l_1,l_2))=T$ is used. Reading a field $F$ from an
object is written $(T,l_1,l_2)_F$ which either is $l_1$ if $F=L$ or $l_2$
if $F=R$. To set field $F$ to a new location $l$, we use the notation
$(T,l_1,l_2)!_Fl$, which yields a new triple $(T,l,l_2)$ if $F=L$ or a new
triple $(T,l_1,l)$ if $F=R$.

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment $E$ and a heap $H$, which may be changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would have
been malformed. The environment could not directly map variables to objects,
because several variables can point to the \emph{same} object, because of aliasing.

We use the following notation for updating partial functions:
$E[v\mapsto l]$ denotes the environment which is just like $E$, but maps $v$ to
$l$.

The \lstinline{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
heap under a fresh location $l$ and adds the result variable to the environment,
mapping it to the new location $l$.

The \lstinline{get} operation reads a field $F$ out of an object, and adds the result
variable to the environment, mapping it to the read location. The heap is
unchanged.

The \lstinline{set} operation changes field $F$ of an object stored at the location that
variable $v$ maps to. The new value of the field is the location in variable
$u$. The environment is unchanged.

The \lstinline{guard_class} operation is used to check whether the object stored at the location
that variable $v$ maps to is of type $T$. If that is the case, then execution
continues without changing heap and environment. Otherwise, execution is
stopped.

% subsection Operational Semantics of Traces (end)

\subsection{Optimizing Traces}
\label{sub:formalopt}

>>>>

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{setspace}
\usepackage{listings}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Subtitle Text, if any}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

The text of the paper begins here.

@@

def f(y):
   step = BoxedInteger(-1)
   while y.is_positive():
      y = y.add(step)
   return res
\end{lstlisting}
\caption{An ``Interpreter'' for a Tiny Dynamic Language Written in RPython}
\label{fig:objmodel}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
# arguments to the trace: $p_{0}$, $p_{1}$
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}
